---
title: "Introduction to assetIS"
author: "Samuel Anyaso-Samuel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to assetIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(2025)
```

# Overview

The **assetIS** package provides importance sampling (IS) estimators of
extreme tail probabilities for the
[ASSET](https://doi.org/10.1016/j.ajhg.2012.03.015) subset-based
meta-analysis statistic (Bhattacharjee et al., 2012). The ASSET statistic
is
$$Z_{\text{ASSET}} = \max_{A \neq \varnothing} |Z_A|,$$
where the maximum is taken over all $2^M - 1$ non-empty subsets $A$ of
$M$ studies (or cell types). Computing the significance of
$Z_{\text{ASSET}}$ requires evaluating an extreme tail probability, which
is computationally prohibitive by naive Monte Carlo when the p-value is
very small.

IS solves this by exponentially tilting the simulation distribution toward
the region of interest, achieving substantial variance reductions. The
package implements four estimators:

| Setting | Independent | Correlated |
|---|---|---|
| Large-sample (Gaussian) | `asset_is_pvalue_ind()` | `asset_is_pvalue_corr()` |
| Small-sample (non-Gaussian) | `asset_is_nonnorm_ind()` | `asset_is_nonnorm_corr()` |

```{r load}
library(assetIS)
```

# Gaussian IS: Independent Studies

## Setup

Consider $M = 5$ independent studies with moderate sample sizes. The
Gaussian IS is appropriate when each $z_m \sim N(0,1)$ under the null.

```{r gaussian_ind_setup}
set.seed(42)
M   <- 5
n   <- c(500, 600, 450, 700, 550)     # study sample sizes
z   <- c(3.1, 2.9, 0.2, -0.1, 0.4)   # observed score statistics

## Observed ASSET statistic
b_obs <- max(abs(z))
cat("Observed ASSET statistic:", round(b_obs, 3), "\n")
```

## Single threshold

```{r gaussian_ind_single}
res_single <- asset_is_pvalue_ind(
  z         = z,
  n         = n,
  b         = b_obs,
  K         = 5000L,
  two.sided = TRUE,
  seed      = 1L
)
print(res_single)
```

## Range of thresholds

When `b` is a vector, a single tilting parameter $\xi = \text{median}(b)$
is reused, making the call no more expensive than a single threshold.

```{r gaussian_ind_range}
b_grid <- seq(2.5, 5.5, by = 0.25)

res_range <- asset_is_pvalue_ind(
  z         = z,
  n         = n,
  b         = b_grid,
  K         = 1e4L,
  two.sided = TRUE,
  seed      = 2L
)
head(res_range)
```

```{r gaussian_ind_plot, fig.cap="IS p-value estimates over a grid of thresholds (log scale). Shaded band is a 95% confidence interval."}
plot(res_range$b, log10(res_range$p.est),
     type = "l", lwd = 2,
     xlab = "Threshold b", ylab = expression(log[10](hat(p))),
     main = "Gaussian IS (independent studies)")
polygon(c(res_range$b, rev(res_range$b)),
        log10(c(res_range$p.upper, rev(res_range$p.lower))),
        col = adjustcolor("steelblue", 0.2), border = NA)
lines(res_range$b, log10(res_range$p.lower), lty = 2, col = "steelblue")
lines(res_range$b, log10(res_range$p.upper), lty = 2, col = "steelblue")
```

## Case-control studies

For case-control data, replace $n_m$ with the effective sample size
$n_m^{\text{eff}} = n_m \phi_m(1-\phi_m)$, where $\phi_m$ is the case
fraction in study $m$.

```{r gaussian_ind_cc}
phi   <- c(0.30, 0.40, 0.50, 0.35, 0.45)
n_eff <- n * phi * (1 - phi)

asset_is_pvalue_ind(
  z         = z,
  n         = n_eff,
  b         = b_obs,
  K         = 5000L,
  two.sided = TRUE,
  seed      = 3L
)
```

# Gaussian IS: Correlated Studies

When studies share subjects, the score statistics are correlated with
known $M \times M$ correlation matrix $\Sigma$.

```{r gaussian_corr}
rho   <- 0.35
Sigma <- outer(seq_len(M), seq_len(M), function(i, j) rho^abs(i - j))

res_corr <- asset_is_pvalue_corr(
  z         = z,
  n         = n,
  Sigma     = Sigma,
  b         = b_obs,
  K         = 5000L,
  two.sided = TRUE,
  seed      = 4L
)
print(res_corr)
```

# Non-Gaussian IS: Independent Cell Types

When sample sizes are small (e.g., single-cell eQTL, rare-disease
meta-analysis), the Gaussian approximation fails. The non-Gaussian IS
operates conditional on the observed phenotype matrix $Y$, simulating
genotypes from a discrete Hardy--Weinberg distribution.

## eQTL setting

We consider $C = 5$ cell types, $N = 100$ subjects, and MAF $f = 0.15$.

```{r nonnorm_ind_eqtl}
set.seed(42)
N <- 100; C <- 5; f <- 0.15

## Centered and scaled expression matrix
Y <- scale(matrix(rnorm(N * C), N, C))

## Simulate a genotype vector and compute the ASSET statistic
g       <- rbinom(N, 2, f)
sigma_g <- sqrt(2 * f * (1 - f))
z_eqtl  <- as.numeric(crossprod(Y, g - 2 * f)) / (sqrt(N) * sigma_g)
b_eqtl  <- max(abs(z_eqtl))
cat("Observed ASSET (eQTL):", round(b_eqtl, 3), "\n")

res_eqtl <- asset_is_nonnorm_ind(
  Y         = Y,
  f         = f,
  b         = b_eqtl,
  K         = 2000L,
  two.sided = TRUE,
  seed      = 5L
)
print(res_eqtl)
```

## Case-control meta-analysis (same ancestry)

```{r nonnorm_ind_cc}
set.seed(7)
N_total  <- 300
study_id <- sample(seq_len(C), N_total, replace = TRUE)
phi_c    <- runif(C, 0.25, 0.40)
y_bin    <- rbinom(N_total, 1, prob = phi_c[study_id])

## Build centered outcome matrix
Ycc <- matrix(0, N_total, C)
for (i in seq_len(N_total))
  Ycc[i, study_id[i]] <- y_bin[i] - phi_c[study_id[i]]

## Compute score statistics using common MAF f = 0.2
f_cc    <- 0.20
sigma_g <- sqrt(2 * f_cc * (1 - f_cc))
g_cc    <- rbinom(N_total, 2, f_cc)
z_cc    <- vapply(seq_len(C), function(c) {
  idx <- which(study_id == c)
  Nc  <- length(idx)
  sum(Ycc[idx, c] * (g_cc[idx] - 2 * f_cc)) /
    (sqrt(Nc * phi_c[c] * (1 - phi_c[c])) * sigma_g)
}, numeric(1L))
b_cc <- max(abs(z_cc))

res_cc <- asset_is_nonnorm_ind(
  Y         = Ycc,
  f         = f_cc,
  b         = b_cc,
  study_id  = study_id,
  K         = 2000L,
  two.sided = TRUE,
  seed      = 8L
)
print(res_cc)
```

## Multi-ancestry case-control (study-specific MAFs)

Pass a length-$C$ vector `f` to handle different allele frequencies
across ancestries.

```{r nonnorm_ind_multiancestry}
f_multi <- c(0.15, 0.22, 0.10, 0.30, 0.18)

res_multi <- asset_is_nonnorm_ind(
  Y         = Ycc,
  f         = f_multi,
  b         = b_cc,
  study_id  = study_id,
  K         = 2000L,
  two.sided = TRUE,
  seed      = 9L
)
print(res_multi)
```

# Non-Gaussian IS: Correlated Cell Types

When gene expression is correlated across cell types, the GLS-based
estimator accounts for the empirical correlation
$\widehat\Sigma_Y = Y^\top Y / N$.

```{r nonnorm_corr}
set.seed(42)
N <- 150; C <- 5; f <- 0.15; rho <- 0.50

## Correlated expression data
if (requireNamespace("MASS", quietly = TRUE)) {
  Sigma_expr <- (1 - rho) * diag(C) + rho
  Y_corr     <- MASS::mvrnorm(N, mu = rep(0, C), Sigma = Sigma_expr)
} else {
  ## fallback: independent columns
  Y_corr <- matrix(rnorm(N * C), N, C)
}
Y_corr <- scale(Y_corr)

g      <- rbinom(N, 2, f)
sigma_g <- sqrt(2 * f * (1 - f))

## GLS scores (approximate, using empirical Sigma_z)
Sigma_z_hat <- crossprod(Y_corr) / N
diag(Sigma_z_hat) <- 1
z_corr <- as.numeric(crossprod(Y_corr, g - 2 * f)) / (sqrt(N) * sigma_g)
b_corr <- max(abs(z_corr))

res_nncorr <- asset_is_nonnorm_corr(
  Y         = Y_corr,
  f         = f,
  b         = b_corr,
  K         = 2000L,
  two.sided = TRUE,
  scale_Y   = FALSE,   # already scaled
  seed      = 10L
)
print(res_nncorr)
```

# Parallel Computation

All four functions support parallel execution via `doParallel`. Set
`parallel = TRUE` and optionally specify `n.cores`.

```{r parallel, eval=FALSE}
## Not run in vignette (requires multiple cores)
res_par <- asset_is_pvalue_ind(
  z         = z,
  n         = n,
  b         = b_grid,
  K         = 1e5L,
  two.sided = TRUE,
  parallel  = TRUE,
  n.cores   = 4L,
  seed      = 99L
)
```

# Choosing the Number of Simulations

The table below gives approximate guidance on the number of simulations
$K$ needed to achieve a relative error of roughly 10%.

| Target p-value | Naive MC | IS (approximate) |
|---|---|---|
| $10^{-3}$ | $10^5$ | $5{,}000$ |
| $10^{-5}$ | $10^7$ | $10{,}000$ |
| $10^{-7}$ | $10^9$ | $50{,}000$ |

For genome-wide significance ($p \approx 5 \times 10^{-8}$) we recommend
$K \geq 10^5$ for the Gaussian estimators and $K \geq 5 \times 10^4$ for
the non-Gaussian estimators with modest $C$ ($\leq 10$).

# Session Info

```{r session}
sessionInfo()
```

# References

Bhattacharjee S, Rajaraman P, Bhattacharjee S, et al. (2012). A
subset-based approach improves power and interpretation for the combined
analysis of genetic association studies of heterogeneous traits.
*American Journal of Human Genetics*, **90**(5), 821--835.

Shi J and Siegmund D (2007). Importance sampling for estimating p-values
in linkage analysis. *Journal of the American Statistical Association*, **102**(479), 929--937.

Siegmund D (1976). Importance sampling in the Monte Carlo study of
sequential tests. *The Annals of Statistics*, **4**(4), 673--684.
